---
title: "Modification Indices in umx"
subtitle: "Exploring model improvement suggestions with umxMI()"
author: "Timothy C. Bates"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Modification Indices in umx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
library(umx)
library(OpenMx)
```

## Introduction

**Modification indices** (MIs) flag single parameters whose addition or removal would most improve model fit (if freed and the model re-estimated). They are useful for exploring *where* a misspecified model struggles, but they are **not** a substitute for theory-driven model comparison.

`umxMI()` wraps [OpenMx::mxMI()](https://openmx.ssri.psu.edu/) and prints the largest indices (by default those above $\chi^2_1(0.01) = 6.63$). Use `full = FALSE` for a fast screen (only the tested parameter moves); `full = TRUE` is slower but allows all parameters to adjust.

```{r note, echo = FALSE}
# Post-hoc MI fishing is poor practice; use competing substantive models when you can.
```

This vignette walks through the classic Holzinger & Swineford (1939) data ([`HS.ability.data`](https://openmx.ssri.psu.edu/reference/HS.ability.data)), building increasingly adequate models and checking what `umxMI()` suggests at each step.

## Data

The nine tests we use are the standard teaching subset of `HS.ability.data`:

```{r data-setup}
data("HS.ability.data", package = "OpenMx")

manifests = c("visual", "cubes", "flags", "paragrap", "sentence", "wordm",
  "addition", "counting", "straight")

visualItems = c("visual", "cubes", "flags")
verbalItems = c("paragrap", "sentence", "wordm")
speedItems  = c("addition", "counting", "straight")

# Scale for stable optimization (recommended for this dataset in umx)
df = umx_scale(HS.ability.data)
```

## One-factor model

We begin with a single general factor `g` loading on all nine tests.

```{r one-factor, results = "hide", message = FALSE}
m1 = umxRAM("Holzinger_1f", data = df,
  umxPath("g", to = manifests),
  umxPath(v.m. = manifests),
  umxPath(v1m0 = "g")
)
```

```{r one-factor-summary}
umxSummary(m1)
```

Fit is poor (as expected for a single factor on these data). What does `umxMI()` suggest?

```{r one-factor-mi}
umxMI(m1, full = FALSE)
```

Large indices often point to **within-domain residual covariances** (e.g. `flags_with_visual`) or **cross-domain loadings** that a single factor cannot represent. That pattern hints at multiple abilities rather than one `g`.

## Three-factor model

Holzinger and Swineford's data are usually modeled with three correlated abilities. We use the conventional lavaan names `spatial`, `verbal`, and `speed` for the visual, verbal, and speed domains (latent names must differ from manifest names such as `visual`).

```{r three-factor, results = "hide", message = FALSE}
model3f = "
  spatial =~ visual + cubes + flags
  verbal  =~ paragrap + sentence + wordm
  speed   =~ addition + counting + straight
"
m3 = umxRAM(model3f, data = df, name = "Holzinger_3f")
```

```{r three-factor-summary}
umxSummary(m3)
```

```{r three-factor-compare}
umxCompare(m3, m1)
```

The three-factor model fits substantially better than the one-factor model. Remaining MI suggestions are smaller but still highlight residual associations not yet in the model:

```{r three-factor-mi}
umxMI(m3, full = FALSE)
```

## Bifactor model

A bifactor model adds a general factor `g` alongside orthogonal specific factors. Here `g` loads on all measures; `spatial`, `verbal`, and `speed` load only on their domain items and are uncorrelated with each other and with `g`.

```{r bifactor, results = "hide", message = FALSE}
modelBifactor = "
  g =~ visual + cubes + flags + paragrap + sentence + wordm + addition + counting + straight
  spatial =~ visual + cubes + flags
  verbal  =~ paragrap + sentence + wordm
  speed   =~ addition + counting + straight
  spatial ~~ 0*verbal + 0*speed + 0*g
  verbal  ~~ 0*speed + 0*g
  speed   ~~ 0*g
"
mBio = umxRAM(modelBifactor, data = df, name = "Holzinger_bifactor")
```

```{r bifactor-summary}
umxSummary(mBio)
```

```{r bifactor-compare}
umxCompare(mBio, m3)
```

```{r bifactor-mi}
umxMI(mBio, full = FALSE)
```

For this bifactor solution, few (if any) indices exceed the 1% $\chi^2$ threshold — the model is already doing a good job, and remaining suggestions are minor.

## Multi-group bifactor (sex differences)

Finally, we fit the bifactor model separately by sex using the `Gender` column in `HS.ability.data` (`Female` / `Male`).

```{r multi-group-fit, results = "hide", message = FALSE}
mMG = umxRAM(modelBifactor, data = df, name = "Holzinger_bifactor_MG", group = "Gender")
```

```{r multi-group-summary}
umxSummary(mMG)
```

Multi-group modification indices require **recent OpenMx** support inside `mxMI()`. Older releases errored on multi-group models. The chunk below is wrapped so the vignette still builds on CRAN if your installed OpenMx predates that support:

```{r multi-group-mi, error = TRUE}
mgMi = tryCatch(
  umxMI(mMG, full = FALSE),
  error = function(e) {
    message(
      "Multi-group umxMI() is not available with this OpenMx build.\n",
      "Error was: ", conditionMessage(e), "\n",
      "Install a current OpenMx release when available."
    )
    NULL
  }
)
mgMi
```

When it runs successfully, parameter labels include group suffixes (e.g. `_Female`), highlighting paths whose addition or release might improve fit **in one sex** or **differentially** across groups.

## Summary

| Model | Role | Typical `umxMI()` story |
|-------|------|-------------------------|
| One-factor | Baseline | Large cross-domain / within-domain suggestions |
| Three-factor | Substantive CFA | Better fit; smaller residual suggestions |
| Bifactor | General + specific | Often few large indices when well specified |
| Multi-group bifactor | Sex differences | Group-specific suggestions (needs current OpenMx) |

**Take-home:** Use `umxMI()` to diagnose *where* a model misfits; prefer `umxCompare()` on **theory-driven** competing models for inference.

## References

Holzinger, K. J., & Swineford, F. (1939). A study in factor analysis: The stability of a bi-factor solution. *Supplementary Educational Monographs*, 48. University of Chicago Press.