---
title: "Uncertainty in SEM"
subtitle: "Standard Errors, Robust SEs (Maximum Likelihood Robust) and and Profile Likelihood CIs in umx"
author: "Timothy C. Bates"
date: "`r Sys.Date()`"
output: 
	rmarkdown::html_vignette:
	theme: darkly          # or flatly, slate, etc. (Bootswatch themes)
	highlight: zenburn     # or breezedark, tango, pygments, kate, espresso, etc.
	toc: true
	toc_float: true
	code_folding: hide     # optional nice-to-have
	# css: custom.css        # optional, for further tweaks
vignette: >
  %\VignetteIndexEntry{Confidence Intervals and Standard Errors in SEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
  knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
library(umx)
library(OpenMx)
```

## Introduction

Fitting a Structural Equation Model (SEM) yields **point estimates** (e.g., a path coefficient of $0.45$ or a variance component of $0.21$). However, scientific inference requires representing the **uncertainty** of these estimates. Moreover, we are interested in the fit of the model itself to the data.

`umx` (+ `OpenMx`) provides three primary ways to assess uncertainty:
1. **Standard Maximum Likelihood Standard Errors** (Wald-based).
2. **Robust Standard Errors** (Huber-White / "Sandwich" estimator).
3. **Profile Likelihood Confidence Intervals** (likelihood-based).

This vignette explains the differences between these methods, when to use each, and how to implement them in your `umx` workflow.

---

## 1. Point Estimates vs. Inference (SEs vs. CIs)

### Standard Errors (Wald CIs)
Standard errors represent the standard deviation of a parameter's sampling distribution. By default, packages construct **Wald confidence intervals** using the standard error:
$$\hat{\theta} \pm z_{\alpha/2} \times \text{SE}$$

* **Assumption:** The sampling distribution of the estimator is symmetric and normal.
* **Limitations:** This assumption breaks down in several common scenarios:
  * Small sample sizes.
  * Parameters near boundaries (e.g., variances, which are bounded at 0).
  * Highly non-linear parameter functions (e.g., indirect effects in mediation models).

### Profile Likelihood Confidence Intervals
Instead of assuming a normal distribution, **Profile Likelihood Confidence Intervals** construct the interval directly from the log-likelihood surface. 

To find the lower boundary of a 95% profile CI for a parameter $\theta$:
1. The parameter $\theta$ is fixed to a value below its maximum likelihood estimate (MLE).
2. All other parameters in the model are re-optimized to find the maximum possible log-likelihood.
3. Steps 1-2 are repeated until we find the point where the log-likelihood drops by exactly $\chi^2_1(0.95)/2 \approx 1.92$ units compared to the MLE.

Because they re-optimize the model at each step, profile CIs:
* Respect the asymmetry of the parameter space (e.g., a CI for variance might be $[0.02, 0.45]$ rather than $[-0.10, 0.35]$).
* Do not rely on local curvature (Hessian matrix) at the MLE.
* Restrict values to boundary constraints.

---

## 2. ML Standard Errors vs. Robust Standard Errors

### Standard ML SEs
Standard Maximum Likelihood standard errors are derived from the local curvature of the log-likelihood surface at the peak (the inverse of the Hessian matrix / information matrix).
* **When to use:** When the data are multivariate normal, the model is correctly specified, and the sample size is large.

### Robust Standard Errors (Sandwich Estimator)
When data are non-normal, standard ML standard errors are typically underestimated, leading to inflated Type-I error rates. 
**Robust standard errors** (also called Huber-White or "sandwich" SEs) correct for this.

The robust covariance matrix is calculated as:
$$\text{Cov}_{\text{robust}} = H^{-1} B H^{-1}$$
* **$H$ (The "Bread"):** The standard Hessian matrix (curvature of the likelihood surface).
* **$B$ (The "Meat"):** The variability of the individual case-wise gradients.

#### When to prefer Robust SEs:
* Mild-to-moderate violation of multivariate normality.
* Minor model miss-specifications.
* Clustered or nested data (if using a cluster-robust correction).

#### Limitations:
* Robust standard errors do not fix biased point estimates resulting from severe misspecification.
* They can be unstable in small samples ($N < 100$).
* They still rely on the Wald symmetry assumption for confidence intervals.

---

## 3. Practical Workflow in `umx`

Let's illustrate these options using the built-in `demoOneFactor` dataset.

### Step 1: Fit a Base Model
We fit a simple one-factor model using `umxRAM()`. By default, `umxSummary()` reports standard ML standard errors.


```{r load_data, message=FALSE, warning=FALSE}
# Load data
data(demoOneFactor)
manifests = names(demoOneFactor)

# Fit model
m1 = umxRAM("One_Factor_Model", data = demoOneFactor, 
	umxPath("g", to = manifests),
	umxPath(v.m. = manifests),
	umxPath(v1m0 = "g")
)
```

We can inspect the standard ML standard errors and Wald CIs:
```{r summary_SE, message=FALSE}
umxSummary(m1, std = TRUE, uncertainty = "SE")
```

### Step 2: Request Robust Standard Errors
To obtain robust standard errors, we can use `uncertainty = "RobustSE"`.

# Compute robust standard errors

```{r robust_Robust, message=FALSE}
umxSummary(m1, std = TRUE, uncertainty = "RobustSE")
```

We can also use `imxRobustSE()` from `OpenMx` on our fitted model, returning a vector of SEs:

```{r imxRobust_se, message=FALSE}
m1Robust = imxRobustSE(m1)
```

# Compute CI
Finally we can extract profile-based confidence intervals (CI). These do not have to be symmetrical.

```{r CI, message=FALSE}
umxSummary(m1, std = TRUE, uncertainty = "CI")

# Extract standard errors to compare
stdSE = m1$output$standardErrors

# Compare SEs side-by-side
compareSE = data.frame(
	Parameter             = row.names(stdSE),
	Standard_ML_SE = round(stdSE, 4),
	Robust_SE           = round(m1Robust, 4)
)
print(compareSE)
```

If the robust standard errors are larger than the standard ML standard errors, it suggests that multivariate non-normality is present in your data.

### Step 3: Compute Profile Likelihood Confidence Intervals
Profile likelihood intervals are requested during model building or via `umxConfint()`. To compute CIs for specific paths (e.g., the loadings of $x1$ and $x2$ on $g$):

```{r profile_ci_specific, message=FALSE, warning=FALSE}
# Request profile likelihood CIs for specific parameters
m1CI = umxConfint(m1, parm = c("g_to_x1", "g_to_x2"), run = TRUE)
```

# View summary showing profile CIs

```{r summary_ci_print, message=FALSE}
umxSummary(m1CI, uncertainty = "CI")
```

Note that the profile CIs are asymmetric and respect the shape of the likelihood surface.

---

## 4. Summary: When to Prefer Which?

| Feature | Standard ML SEs | Robust SEs (Sandwich) | Profile Likelihood CIs |
| :--- | :--- | :--- | :--- |
| **Computational Cost** | Extremely low (one Hessian inversion) | Low (gradient calculations) | High (requires re-optimization) |
| **Normality Assumption** | Strict | Relaxed | Strict (unless robust version used) |
| **Handles Boundaries?** | No (can yield negative variances) | No | Yes (respects constraints) |
| **Asymmetric CIs?** | No (always symmetric) | No | Yes (reflects log-likelihood shape) |
| **Best For...** | Large samples, normal data | Non-normal data, rapid estimation | Key parameters, small samples, bounded parameters |

### Recommendations for Reporting
1. **Use Robust SEs** for general parameters (such as minor residual variances or covariances) if your data shows signs of non-normality.
2. **Compute Profile Likelihood CIs** (using `umxConfint`) for your key structural parameters (e.g., hypothesized paths, mediation effects) to ensure your scientific inferences are robust to asymmetry and boundary conditions.
