---
title: "Getting Started with umx"
subtitle: "Your first structural equation model using umxRAM() and umxSummary()"
author: "Timothy C. Bates"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with umx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
library(umx)
```

## Introduction

`umx` is a package that makes building, running, and reporting Structural Equation Models (SEM) fast and readable in R.

This vignette shows a new user how to:

1. Load classic data (Holzinger & Swineford, 1939)
2. Build a simple one-factor model using `umxRAM()`
3. View and interpret results using `umxSummary()`

## The Holzinger and Swineford (1939) data

This well-known dataset contains scores on nine cognitive tests for 301 school children. The variables `x1` to `x9` measure different aspects of cognitive ability.

Because `umx` imports `lavaan`, the data is easy to load:

```{r load-data}
# Load the classic Holzinger & Swineford data (x1:x9)
data("HolzingerSwineford1939", package = "lavaan")

# Define the manifest (observed) variables
manifests <- paste0("x", 1:9)

# Quick look at the correlations
round(cor(HolzingerSwineford1939[, manifests]), 2)
```

## Build a one-factor model with `umxRAM()`

`umxRAM()` lets you describe models using simple `umxPath()` statements. Any variable name not in the data is treated as a latent variable.

Here we build a model in which a single latent factor `G` (general ability) accounts for the covariances among x1 to x9.

```{r build-model, results = "hide", message = FALSE}
m1 <- umxRAM("Holzinger_1f", 
  data = HolzingerSwineford1939,
  
  # Factor loadings (G predicts each x)
  umxPath("G", to = manifests),
  
  # Residual variances for the observed variables
  umxPath(var = manifests),
  
  # Fix the latent factor variance to 1 (for identification)
  umxPath(var = "G", fixedAt = 1)
)
```

That's it — `umxRAM()` automatically:

- Detects manifests vs latents
- Adds starting values
- Runs the model
- Produces an initial summary

## View results with `umxSummary()`

`umxSummary()` gives clean, publication-ready output.

```{r summary-basic}
umxSummary(m1)
```

### Standardized solution

Add `std = TRUE` to see standardized estimates (very useful for interpretation):

```{r summary-std}
umxSummary(m1, std = TRUE)
```

The output includes:

- A table of standardized loadings, variances, and means
- Model fit information ($\chi^2$, CFI, TLI, RMSEA)
- Helpful warnings when fit is poor

> Note: A single-factor model is a poor fit to these data (as expected — the data were designed to measure three correlated abilities). This is intentional for demonstration: you can immediately see how `umxSummary()` highlights areas for model improvement.

## Alternative: lavaan-style syntax

`umxRAM()` also accepts lavaan model syntax directly:

```{r lavaan-string, results = "hide", message = FALSE}
m2 <- umxRAM("G =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9", 
             data = HolzingerSwineford1939)
```

```{r lavaan-summary}
umxSummary(m2, std = TRUE)
```

This is handy when you already have a lavaan model string.

## Plotting the model

`umxRAM()` automatically plots the model when you run it interactively. You can also plot manually:

```{r plot, fig.cap="Standardized one-factor Holzinger model"}
plot(m1, std = TRUE)
```

## Next steps

- Learn more about path specification: `?umxPath`
- See all the options for `umxRAM()`: `?umxRAM`
- Control output format: `?umxSummary`
- Compare models: `?umxCompare`
- For twin models see the twin modeling functions (`umxACE()`, etc.)

Welcome to `umx`!

## References

Holzinger, K. J., & Swineford, F. (1939). A study in factor analysis: The stability of a bi-factor solution. *Supplementary Educational Monographs*, 48. University of Chicago Press.
```

The content is written.

I should also update DESCRIPTION to declare the VignetteBuilder.

This is necessary for the vignette to be recognized.