---
title: "Weighted Least Squares (WLS) in umx"
subtitle: "Fitting and Interpreting Ordinal Factor Models"
author: "Timothy C. Bates"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Weighted Least Squares (WLS) in umx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
library(umx)
library(OpenMx)
library(lavaan)
```

## Introduction

Most Structural Equation Models (SEMs) are fitted using **Maximum Likelihood (ML)**, which assumes that the manifest variables are continuous and multivariate normal. However, in behavioral genetics, psychology, and social sciences, variables are frequently categorical or ordinal (e.g., binary yes/no, or Likert scales).

When manifest variables are ordinal, standard ML yields biased parameter estimates and severely underestimated standard errors. The standard solution is **Weighted Least Squares (WLS)**, which estimates thresholds for the ordinal indicators and fits the model to the resulting polychoric/tetrachoric correlation matrix.

This vignette demonstrates how to fit and interpret a 1-factor model under different estimators (ML, full WLS, and Diagonally Weighted Least Squares) using the built-in Holzinger-Swineford ordinal dataset (`HSwls`) in `umx`.

---

## 1. The Holzinger-Swineford Ordinal Dataset (`HSwls`)

The `HSwls` dataset contains cognitive test scores `x1` to `x9` for 301 students from two schools. In this version of the data, the variables `x1` to `x9` have been converted into ordered factors (with 2, 3, or 4 categories) using quantile-based cuts:

* `x1` to `x3` represent visual ability (binary / 2 levels).
* `x4` to `x6` represent verbal ability (3 levels).
* `x7` to `x9` represent speed (4 levels).

```{r load_data}
data(HSwls, package = "umx")
str(HSwls[, paste0("x", 1:9)])
```

---

## 2. Fitting a 1-Factor Model under Three Estimators

We will fit a 1-factor model (`g =~ x1 + ... + x9`) using three different estimators:
1. **Maximum Likelihood (ML)**: Treats the ordinal indicators as continuous (incorrect but useful for comparison).
2. **Full Weighted Least Squares (WLS)**: Uses the full asymptotic covariance matrix of thresholds/correlations.
3. **Diagonally Weighted Least Squares (DWLS / WLSMV)**: Uses only the diagonal of the weight matrix (standard for modest sample sizes).

In `umxRAM`, the estimator is requested using the `type` argument:
* `type = "FIML"` (or `"Auto"` for continuous data) for ML.
* `type = "WLS"` for Full WLS.
* `type = "DWLS"` for Diagonally Weighted Least Squares.

### 1. Maximum Likelihood (ML)
First, we fit using ML by converting variables to numeric (to treat them as continuous):

```{r fit_ml, message=FALSE, warning=FALSE}
# Convert ordinal factors to numeric for ML estimation
dfMl = HSwls
dfMl[, paste0("x", 1:9)] = lapply(dfMl[, paste0("x", 1:9)], as.numeric)

mML = umxRAM("One_Factor_ML", data = dfMl, type = "FIML",
	umxPath("g", to = paste0("x", 1:9)),
	umxPath(var = paste0("x", 1:9)),
	umxPath(var = "g", fixedAt = 1)
)
```

### 2. Full Weighted Least Squares (WLS)
Next, we fit using the full WLS estimator on the original ordinal factors:

```{r fit_wls, message=FALSE, warning=FALSE}
mWLS = suppressWarnings(umxRAM("One_Factor_WLS", data = HSwls, type = "WLS",
	umxPath("g", to = paste0("x", 1:9)),
	umxPath(var = paste0("x", 1:9)),
	umxPath(var = "g", fixedAt = 1)
))
```

### 3. Diagonally Weighted Least Squares (DWLS)
Finally, we fit using the DWLS estimator:

```{r fit_dwls, message=FALSE, warning=FALSE}
mDWLS = suppressWarnings(umxRAM("One_Factor_DWLS", data = HSwls, type = "DWLS",
	umxPath("g", to = paste0("x", 1:9)),
	umxPath(var = paste0("x", 1:9)),
	umxPath(var = "g", fixedAt = 1)
))
```

---

## 3. Comparing Estimates and Standard Errors

Let's extract and compare the standardized factor loadings across the three models:

```{r compare_loadings, message=FALSE}
# Extract standardized coefficients
coefML = coef(mML)
coefWLS = coef(mWLS)
coefDWLS = coef(mDWLS)

# Loadings list (paths from g to x1:x9)
loadings = paste0("g_to_x", 1:9)

# Align into a summary table
compareTable = data.frame(
	Path = loadings,
	ML_Estimate = round(coefML[loadings], 3),
	WLS_Estimate = round(coefWLS[loadings], 3),
	DWLS_Estimate = round(coefDWLS[loadings], 3)
)
print(compareTable)
```

### Key Interpretations:
* **Underestimated ML Loadings:** Treating ordinal indicators as continuous (ML) typically underestimates factor loadings because the non-linear categorization threshold is ignored.
* **WLS vs. DWLS:** Both WLS and DWLS recover the true underlying latent relationships, but DWLS is often more stable and exhibits lower mean squared error in modest sample sizes ($N < 1000$).

---

## 4. Model Fit and Comparison under WLS

### Fit Indices
Conventional fit indices (CFI, TLI, RMSEA) are calculated using $\chi^2$. In WLS/DWLS, the discrepancy function is inflated by the weight matrix, causing standard fit index cutoffs (e.g., CFI $> 0.95$, RMSEA $< 0.06$) to be **invalid**. 

When running `umxSummary()` on a WLS model, `umx` automatically uses **Satorra-Bentler (2010)** robust corrections to adjust fit statistics:

```{r summary_wls, message=FALSE}
umxSummary(mDWLS, std = TRUE, showEstimates = "std")
```

### Model Comparison
When comparing nested WLS/DWLS models, we cannot simply subtract the raw chi-squares because the weight matrices differ between the models. `umxCompare()` automatically performs a **Strict Satorra-Bentler scaled difference test**:

```{r nested_compare, message=FALSE, warning=FALSE}
# Fit nested model (drop g to x1 loading)
mDWLS_nested = suppressWarnings(umxModify(mDWLS, update = "g_to_x1", name = "Nested_DWLS"))

# Compare models
umxCompare(mDWLS, mDWLS_nested)
```

---

## 5. Practical Considerations

### 1. Sample Size Requirements
* **Full WLS:** Requires a very large sample size ($N > 1000$, and often $N > 2000$ for models with many indicators). This is because the full asymptotic covariance matrix ($V$) grows rapidly with the number of thresholds, requiring a huge amount of data to estimate accurately. If $N$ is too small, $V$ becomes non-positive definite or highly unstable.
* **DWLS:** Only uses the diagonal elements of $V$, meaning it requires much smaller sample sizes ($N > 200$) to achieve stable and accurate parameter estimates.

### 2. Convergence Issues
If a WLS model fails to converge:
* Check for **empty threshold categories** (zero counts in a cell). If a category has zero observations, the threshold cannot be estimated. You may need to collapse adjacent categories.
* Switch to **DWLS** if you are currently using full WLS.

### 3. Reporting Guidelines
When reporting WLS results in journals:
1. Specify which estimator was used (e.g., "Diagonally Weighted Least Squares (DWLS) with robust standard errors").
2. Report the **Satorra-Bentler robust $\chi^2$** and degrees of freedom.
3. Report **SRMR** as the primary index of absolute fit (cutoff $< 0.08$ or $< 0.10$).
4. Emphasize nested model comparison results using the **Satorra-Bentler difference test** (from `umxCompare`).
