---
title: "Weighted Least Squares (WLS) in umx"
subtitle: "Fitting and Interpreting Ordinal Factor Models"
author: "Timothy C. Bates"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Weighted Least Squares (WLS) in umx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
library(umx)
library(OpenMx)
umx_set_auto_plot(FALSE)
```

## Introduction

Most Structural Equation Models (SEMs) are fitted using **Maximum Likelihood (ML)**, which assumes that the manifest variables are continuous and multivariate normal. However, in behavioral genetics, psychology, and social sciences, variables are frequently categorical or ordinal (e.g., binary yes/no, or Likert scales).

When manifest variables are ordinal, standard ML yields biased parameter estimates and severely underestimated standard errors. The standard solution is **Weighted Least Squares (WLS)**, which estimates thresholds for the ordinal indicators and fits the model to the resulting polychoric/tetrachoric correlation matrix.

This vignette demonstrates how to fit and interpret a 1-factor model under different estimators (ML, full WLS, and Diagonally Weighted Least Squares) using the built-in Holzinger-Swineford ordinal dataset (`HSwls`) in `umx`.

---

## 1. The Holzinger-Swineford Ordinal Dataset (`HSwls`)

The `HSwls` dataset contains cognitive test scores `x1` to `x9` for 301 students from two schools. In this version of the data, the variables `x1` to `x9` have been converted into ordered factors (with 2, 3, or 4 categories) using quantile-based cuts:

* `x1` to `x3` represent visual ability (binary / 2 levels).
* `x4` to `x6` represent verbal ability (3 levels).
* `x7` to `x9` represent speed (4 levels).

```{r load_data}
data(HSwls, package = "umx")
str(HSwls[, paste0("x", 1:9)])
```

---

## 2. Fitting a 1-Factor Model under Three Estimators

We will fit a 1-factor model (`g =~ x1 + ... + x9`) using three different estimators:
1. **Maximum Likelihood (ML)**: Treats the ordinal indicators as continuous (incorrect but useful for comparison).
2. **Full Weighted Least Squares (WLS)**: Uses the full asymptotic covariance matrix of thresholds/correlations.
3. **Diagonally Weighted Least Squares (DWLS / WLSMV)**: Uses only the diagonal of the weight matrix (standard for modest sample sizes).

In `umxRAM`, the estimator is requested using the `type` argument:
* `type = "FIML"` (or `"Auto"` for continuous data) for ML.
* `type = "WLS"` for Full WLS.
* `type = "DWLS"` for Diagonally Weighted Least Squares.

### Live demo: DWLS only (package-check friendly)

Full WLS and continuous-ML-on-ordinal demos are slower (polychoric prep + denser weights). For CRAN/check we fit **DWLS** once on a compact 6-indicator, two-factor model. Run the larger 9-variable / full-WLS variants locally with `eval = TRUE` in the optional chunks below.

```{r fit_dwls, message=FALSE, warning=FALSE}
# Six ordinal indicators, two factors (smaller than full 9-variable 1-factor)
mDWLS = suppressWarnings(umxRAM("Two_Factor_DWLS", data = HSwls, type = "DWLS",
	umxPath("visual", to = c("x1", "x2", "x3")),
	umxPath("verbal", to = c("x4", "x5", "x6")),
	umxPath(var = c("x1", "x2", "x3", "x4", "x5", "x6")),
	umxPath(var = c("visual", "verbal"), fixedAt = 1),
	umxPath("visual", with = "verbal"),
	# Binary indicators: means fixed at 0 for identification under DWLS
	umxPath("one", to = c("x1", "x2", "x3"), fixedAt = 0),
	umxPath("one", to = c("x4", "x5", "x6"))
))
umxSummary(mDWLS, std = TRUE, refModels = FALSE)
```

Optional larger demos (not run during package check):

```{r fit_ml_wls_optional, eval = FALSE}
dfMl = HSwls
dfMl[, paste0("x", 1:9)] = lapply(dfMl[, paste0("x", 1:9)], as.numeric)
mML = umxRAM("One_Factor_ML", data = dfMl, type = "FIML",
	umxPath("g", to = paste0("x", 1:9)),
	umxPath(var = paste0("x", 1:9)),
	umxPath(var = "g", fixedAt = 1)
)
mWLS = umxRAM("One_Factor_WLS", data = HSwls, type = "WLS",
	umxPath("g", to = paste0("x", 1:9)),
	umxPath(var = paste0("x", 1:9)),
	umxPath(var = "g", fixedAt = 1)
)
```

### Nested comparison under DWLS

`umxCompare()` uses a strict Satorra–Bentler scaled difference when Jacobians are available (GenomicMx OpenMx). Keep the nested drop small:

```{r nested_compare, message=FALSE, warning=FALSE}
mDWLS_nested = suppressWarnings(umxModify(mDWLS, update = "visual_to_x1", name = "Nested_DWLS"))
umxCompare(mDWLS, mDWLS_nested)
```

---

## 5. Practical Considerations

### 1. Sample Size Requirements
* **Full WLS:** Requires a very large sample size ($N > 1000$, and often $N > 2000$ for models with many indicators). This is because the full asymptotic covariance matrix ($V$) grows rapidly with the number of thresholds, requiring a huge amount of data to estimate accurately. If $N$ is too small, $V$ becomes non-positive definite or highly unstable.
* **DWLS:** Only uses the diagonal elements of $V$, meaning it requires much smaller sample sizes ($N > 200$) to achieve stable and accurate parameter estimates.

### 2. Convergence Issues
If a WLS model fails to converge:
* Check for **empty threshold categories** (zero counts in a cell). If a category has zero observations, the threshold cannot be estimated. You may need to collapse adjacent categories.
* Switch to **DWLS** if you are currently using full WLS.

### 3. Reporting Guidelines
When reporting WLS results in journals:
1. Specify which estimator was used (e.g., "Diagonally Weighted Least Squares (DWLS) with robust standard errors").
2. Report the **Satorra-Bentler robust $\chi^2$** and degrees of freedom.
3. Report **SRMR** as the primary index of absolute fit (cutoff $< 0.08$ or $< 0.10$).
4. Emphasize nested model comparison results using the **Satorra-Bentler difference test** (from `umxCompare`).
